We Are All Trying to Describe Something And We Don't Have the Words

### 2023-04-20 outputs/paper/paper.qmd
Start cleaning up paper

@@

Sometime in early October of 2022 there was a chemical shift in my brain and I couldn't stop listening to Taylor Swift. I couldn't listen to anything other than Fearless or Red when deadlifting, walking to a friend's house, folding laundry, etc. I listened to so much Taylor in that month alone that she ended up as my #4 artist on Spotify Wrapped, which stops collecting its data for the year on October 31st. It didn't even capture what happened to me after Midnights dropped in November.

@@

Midnights was produced by Swift and her longtime collaborator Jack Antonoff, who is the subject of this paper, and, I suspect, the reason behind both my glitching, and the stranger's "Exactly".

@@

I spoke to Caleb Gamman, a Youtuber and Antonoff-production-style hater who had a tweet go viral around the time of Taylor Swift's *Midnights* release, in which he predicts whether or not a song was produced by Antonoff within seconds. A real-life logistic regression, if you will. Using an actual logistic regression model, I was able to determine that Spotify's `danceability` feature was the most important in predicting whether or not a song had any Antonoff involvement. During an interview with Glen McDonald, an engineer at Spotify, I learned that the `valence` and `danceability` variables were created by hiring college interns and having them tag whether or not songs were positive or danceable. In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness. Against the odds, we can capture something about Antonoff's music with these variables, and it might just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe.

@@

I reached out to Gamman to ask him to explain Antonoff to me, only to find out, on the morning of the interview, that Gamman had released a 29 minute video essay on his Youtube Channel covering [@citeGamman]:

@@

    > ... I got into drumming at the age of like six, so maybe as a drummer the drum samples are \[offensive?\]. I was anachronistically raised on cassettes so, you know maybe I like the sound of audio and maybe that explains why I tend to like dissonance and inharmonic sound, interplay of elements, distortion noise, the clashing of elements. I imagine I've listened to more SoundCloud in my life than radio so uh, maybe I only want to hear amateur internet losers. I'vel iked over 10 000 songs on my Spotify account and I can identify lot of them off the bat, but I'm not that great at identifying music generally, so it's just a compartmentalized media consumption habit. I first did music production in primary school. I made mashups and dubstep. As a teenager I extensively used Juno 6 emulators to make embarrassing synthwave. \[Admittedly that's probably not the subconscious association that he's going for.\] Lyrics aren't of much interest to me, it's just a sound in the mix, and so maybe I don't like the focus on vocals. I've previously described \[my reaction\] as a fight or flight sort of thing, so maybe it's literally is tapping into some memory of trauma. I have an unusually small ear canal so maybe I'm just hearing the high end more acutely. I was hired as a teenager to play drums on a lot of Jack Antonoff songs for a university glee club so maybe I don't like it because of the embarrassment of having ever been associated with a glee club. I liked the album 1989 but to make the song Out of the Woods palettable to me I had to mash it up with M83Midnight City and that got a DMCA takedown so maybe it's a personal vendetta. And of course I have known andd o know people who like this music so maybe I secretly hate my friends and family but I don't know if any of that tells us anything ...

@@

> **Michaela Drouillard:** I know you cover this in the video. I'm sorry. What makes a Jack Antonoff song a Jack Antonoff song?
>
> **Caleb Gamman:** I suppose I do have much more detail that I can go into, but it is always, for me, just a knee jerk. It's not any principled thing. I do have some loose experience with like, audio production stuff. So if I really dig into it, I can maybe point some things out. In the treble mix of the song, like in the very upper end, like sort of above where the vocals are. He often really crushes that down. Like it's lower in volume than anything else, which is sort of a weird effect. Like, when you hear someone speaking, you hear a lot of that, like, noise in their voice. When you yourself are speaking, you hear less of it, right? You hear more of your own voice bouncing around in your head. And so it sort of creates this effect of -- **like he mixes his vocals the way it would sound to the person who's singing them. Which is sort of strange.** He often has no vocals on that upper end, which is very unusual, and then often there are random little bits of noise happening up there which is the sort of thing her likes to do.
>
> So I don't know, like I personally am quite a fan of messier music, you know, more experimental stuff, stuff that has a lot of clashing, sounds, a lot of different interesting pieces coming together. Jack Antonoff does a very good job of making stuff very clean. So you know, all of the notes that are happening simulaneously are perfectly stacked. They're like mathematically stacked .. Like, you have the harmonic spectrum where everything is like a multiple on top of each other. So that's like your notes on a keyboard. And if you played a keyboard note that was a little bit off where it's supposed to be, you're gonna create this interesting effect where these two notes sort of clash against each other and they wiggle around in terms of their tone because you've got these two harmonics happening that don't match. And Jack Antonoff avoids that quite strictly.
>
> His drums are low pass, like the vocals, so you don't get like the tinny high end in the drums or whatever because that tinny high end in the drums is like a bunch of random harmonics all over the place. So if you cut that out, you just get the tone of the drum. And then he achieves those effects using the same sounds like literally the same synthesizer on the same drum machine and stuff.
>
> ....
>
> If you used a cscope spectrogram sort of thing to look at a Jack Antonoff song, you're gonna see a lot of big lines. And then if you're listening to a song by someone else, you're gonna see like a bunch of sort of like static wiggling. It's gonna be really jagged because you've got a bunch of harmonics that don't exactly mix with each other, so they're stacking in all sorts of weird different ways.

@@

library(dplyr)

sub_planets<- read.csv(here::here("inputs/data/subset_planets.csv"))
sub_planets |>
  select(artist_name, valence, danceability, track_name) |>
  head() |>
  knitr::kable(
    col.names = c("Artist Name", 
                  "Valence", 
                  "Danceability",
                  "Track Name")
 )

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-bees
#| tbl-cap: Valence and Danceability Scores for Every Performance of Flight of the Bumblee on Spotify

sub_bees<- read.csv(here::here("inputs/data/subset_bumblebee.csv"))
sub_bees |>
   select(artist_name, valence, danceability, album_release_year, track_name) |>
   head(17) |>
   knitr::kable(
     col.names = c("Artist Name", 
                   "Valence",
                   "Danceability",
                   "Album Release Year",
                   "Track Name")
  )

@@

```{r}
#| label: fig-quadvalence
#| fig-cap: Valence and Energy Quadrant
#| echo: false
#| warning: false
library(ggplot2)
df <- read.csv(here::here("inputs/data/df.csv"))

  ggplot(df, aes(x = energy, y = valence)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$valence), color = "gray") +
    theme_minimal()

@@

  ggplot(df, aes(x = energy, y = danceability)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$danceability), color = "gray") +
    theme_minimal()

@@

By comparing cumulative distribution functions of each variable based on whether or not a song was produced by Antonoff, we can see that some of these variables do capture some sort of underlying Jack-iness (@fig-cdf). For instance, while liveness, instrumentalness, and speechiness are relatively similar, we can see clear differences in danceability, energy, loudness, acousticness, and valence.

```{r}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable, 
#| echo: false
#| warning: false
library(tidyverse)
data_reduced <- read.csv(here::here("inputs/data/data_reduced.csv"))

data_reduced |> 
  pivot_longer(cols = c(danceability, energy, loudness, mode,
               speechiness, acousticness, instrumentalness, liveness, valence,
               tempo),
               names_to = "name",
               values_to = "value"
  ) |> 
  ggplot(aes(x = value, color = jack)) +
  stat_ecdf() +
  facet_wrap(vars(name), 
             nrow = 3, 
             ncol = 4,
             scales = "free") +
  theme_minimal() +
  labs(
    x = "Value",
    y = "Proportion"
  )

#even tho there's similrities, there's differences

Another way to gauge the Antonoff-ification of pop music could be to look at how artists' sounds have changed over time, and whether or not there is any correlation with Antonoff. Here is where the necessary "correlation is not causation" caveat comes in: suggesting that Antonoff is the *cause* for any change in artists' sounds would be an overstep. These plots are here to get a glimpse at pop's evolution over time, and where Antonoff fits in that picture. Each point represents a song.

@@

library(lubridate)
library(tidyverse)
df <- read_csv(here::here("inputs/data/df.csv"))
df$jack <- factor(df$jack)
df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = energy, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Energy") +
  facet_wrap( ~ artist_name, ncol = 3)

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = danceability, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Danceability") +
  facet_wrap(~ artist_name, ncol = 3)

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = valence, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Valence") +
  facet_wrap(~ artist_name, ncol = 3)

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = acousticness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Acousticness") +
  facet_wrap(~ artist_name, ncol = 3)

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = instrumentalness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Instrumentalness") +
  facet_wrap(~ artist_name, ncol = 3)

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = tempo, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Tempo") +
  facet_wrap(~ artist_name, ncol = 3)

@@

Now that we understand that there do seem to be some subtle differences in how variables that are and aren't produced by Jack behave, as seen in our cumulative distribution functions, our quadrants, and some minor trends in artists' sounds over time (specifically, `energy`), we can use a logistic regression to see if a model can a) predict whether or not a song was produced by Antonoff based on these variables, and b) tell us which variables were more significant in determining the output.

@@

data_reduced <- read.csv(here::here("inputs/data/data_reduced.csv"))

data_reduced %>%
  group_by(jack) %>%
  summarise(across(-c(tempo, loudness), mean)) %>%
  pivot_longer(cols = -jack) %>%
  ggplot(aes(value, name, fill = factor(jack))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")

Bracketing the major blindspot that there variables provide a single number to represent the dynamics of an entire song, the model has a few other limitations. First of all, this model assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. (The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.)

@@

data_reduced %>%
  count(jack) %>%
  knitr::kable(col.names = c("Is it Jack?", "Number of Songs"))

@@

After some minor pre-processing using the `tidyverse` R package [@citetidyverse]. I followed Julia Silge's tutorial [@citeSilge] to build a model that deals with the imbalanced classes.

@@

jack_vip <- read.csv(here::here("inputs/data/jack_vip.csv"))

jack_vip |>
   knitr::kable(
     col.names = c("Variable", 
                   "Importance",
                   "Sign")
  )

@@

jack_vip %>%
  group_by(Sign) %>%
  slice_max(Importance, n = 15) %>%
  ungroup() %>%
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) + 
  geom_col() +
  facet_wrap(vars(Sign), scales = "free_y") +
  labs(y = NULL) +
  theme(legend.position = "none")
```

Danceability, with a positive importance score of 4.32, had the highest level of importance in determining whether or not a song was produced or co-produced by Jack Antonoff. Instrumentalness comes in second at 2.59, with a negative score (meaning that negative instrumentalness is correlated), and valence comes in third with a negative score at 2.02.

@@

df %>%
  filter(jack == 1) %>%
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |> 
  head(20) |>
  knitr::kable(
     col.names = c("Artist Name", 
                   "Danceability",
                   "Track")
  )

@@

According to this model, though, danceability is the most important feature in predicting whether or not a song was produced by Jack Antonoff. And the accuracy metrics aren't awful -- the accuracy score is 0.743, and the roc_auc (receiver operating characteristic area under the curve) (more fun to say out loud as an abbreviation. Rock Auck), which is a metric better suited for imbalanced datasets, such as our own, is 0.810. This means that although danceability can't show us the inside of a song, something about this score is telling us something about Jack's music.

@@

```{r}
#| echo: false
#| warning: false
df %>%
  filter(jack == 1) %>%
  ggplot(aes(x = artist_name, y = danceability)) +
  geom_boxplot() +
  ggtitle("Boxplot of Danceability by Artist") +
  coord_flip() +
  theme_minimal()
```

```{r}
#| echo: false
#| warning: false
library(dplyr)
library(ggplot2)

df %>%
  filter(jack == 1) %>%
  count(artist_name) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(artist_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Artists with Most Jack = 1 Values", x = "Artist", y = "Count") +
  theme_minimal()

@@

It only makes sense, of course, to hire a bunch of undergrads and get them to tell us whether or not a few thousand songs are visceral, and then build our own visceral-ness metric, and see if we get the same results -- that danceability is just visceral-ness in a trench coat, and that visceral-ness is the characteristic that helps us best understand what makes Jack, Jack.

>>>>

library(tidyverse)
library(lubridate)

@@

Sometime in early October of 2022 there was a chemical shift in my brain, and I couldn't stop listening to Taylor Swift. I couldn't listen to anything other than Fearless or Red when deadlifting, walking to a friend's house, folding laundry, etc. I listened to so much Taylor in that month alone that she ended up as my #4 artist on Spotify Wrapped, which stops collecting its data for the year on October 31st. It didn't even capture what happened to me after Midnights dropped in November.

@@

Midnights was produced by Swift and her long-time collaborator Jack Antonoff, who is the subject of this paper, and, I suspect, the reason behind both my glitching, and the stranger's "Exactly".

@@

I spoke to Caleb Gamman, a Youtuber and Antonoff-production-style hater who had a tweet go viral around the time of Taylor Swift's *Midnights* release, in which he predicts whether a song was produced by Antonoff within seconds. A real-life logistic regression, if you will. Using an actual logistic regression model, I was able to determine that Spotify's `danceability` feature was the most important in predicting whether a song had any Antonoff involvement. During an interview with Glen McDonald, an engineer at Spotify, I learned that the `valence` and `danceability` variables were created by hiring college interns and having them tag whether songs were positive or danceable. In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness. Against the odds, we can capture something about Antonoff's music with these variables, and it might just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe.

@@

I reached out to Gamman to ask him to explain Antonoff to me, only to find out, on the morning of the interview, that Gamman had released a 29 minute video essay on his YouTube Channel covering [@citeGamman]:

@@

    > ... I got into drumming at the age of like six, so maybe as a drummer the drum samples are \[offensive?\]. I was anachronistically raised on cassettes so, you know maybe I like the sound of audio and maybe that explains why I tend to like dissonance and inharmonic sound, interplay of elements, distortion noise, the clashing of elements. I imagine I've listened to more SoundCloud in my life than radio so uh, maybe I only want to hear amateur internet losers. I've liked over 10 000 songs on my Spotify account and I can identify lot of them off the bat, but I'm not that great at identifying music generally, so it's just a compartmentalized media consumption habit. I first did music production in primary school. I made mashups and dubstep. As a teenager I extensively used Juno 6 emulators to make embarrassing synthwave. \[Admittedly that's probably not the subconscious association that he's going for.\] Lyrics aren't of much interest to me, it's just a sound in the mix, and so maybe I don't like the focus on vocals. I've previously described \[my reaction\] as a fight or flight sort of thing, so maybe it's literally is tapping into some memory of trauma. I have an unusually small ear canal so maybe I'm just hearing the high end more acutely. I was hired as a teenager to play drums on a lot of Jack Antonoff songs for a university glee club so maybe I don't like it because of the embarrassment of having ever been associated with a glee club. I liked the album 1989 but to make the song Out of the Woods palatable to me I had to mash it up with M83Midnight City and that got a DMCA takedown so maybe it's a personal vendetta. And of course I have known and do know people who like this music so maybe I secretly hate my friends and family but I don't know if any of that tells us anything ...

@@

> **Michaela Drouillard:** I know you cover this in the video. I'm sorry. What makes a Jack Antonoff song a Jack Antonoff song?
>
> **Caleb Gamman:** I suppose I do have much more detail that I can go into, but it is always, for me, just a knee jerk. It's not any principled thing. I do have some loose experience with like, audio production stuff. So if I really dig into it, I can maybe point some things out. In the treble mix of the song, like in the very upper end, like sort of above where the vocals are. He often really crushes that down. Like it's lower in volume than anything else, which is sort of a weird effect. Like, when you hear someone speaking, you hear a lot of that, like, noise in their voice. When you yourself are speaking, you hear less of it, right? You hear more of your own voice bouncing around in your head. And so it sort of creates this effect of -- **like he mixes his vocals the way it would sound to the person who's singing them. Which is sort of strange.** He often has no vocals on that upper end, which is very unusual, and then often there are random little bits of noise happening up there which is the sort of thing her likes to do.
>
> So I don't know, like I personally am quite a fan of messier music, you know, more experimental stuff, stuff that has a lot of clashing, sounds, a lot of different interesting pieces coming together. Jack Antonoff does a very good job of making stuff very clean. So you know, all of the notes that are happening simultaneously are perfectly stacked. They're like mathematically stacked .. Like, you have the harmonic spectrum where everything is like a multiple on top of each other. So that's like your notes on a keyboard. And if you played a keyboard note that was a little bit off where it's supposed to be, you're gonna create this interesting effect where these two notes sort of clash against each other and they wiggle around in terms of their tone because you've got these two harmonics happening that don't match. And Jack Antonoff avoids that quite strictly.
>
> His drums are low pass, like the vocals, so you don't get like the tinny high end in the drums or whatever because that tinny high end in the drums is like a bunch of random harmonics all over the place. So if you cut that out, you just get the tone of the drum. And then he achieves those effects using the same sounds like literally the same synthesizer on the same drum machine and stuff.
>
> ....
>
> If you used a cscope spectrogram sort of thing to look at a Jack Antonoff song, you're gonna see a lot of big lines. And then if you're listening to a song by someone else, you're gonna see like a bunch of sort of like static wiggling. It's gonna be really jagged because you've got a bunch of harmonics that don't exactly mix with each other, so they're stacking in all sorts of weird different ways.

@@

sub_planets <- read_csv(here::here("inputs/data/subset_planets.csv"))

sub_planets |>
  select(artist_name, valence, danceability, track_name) |>
  head() |>
  knitr::kable(col.names = c("Artist Name",
                             "Valence",
                             "Danceability",
                             "Track Name"))
```

@@

```{r}
#| echo: false
#| warning: false
#| label: tbl-bees
#| tbl-cap: Valence and Danceability Scores for Every Performance of Flight of the Bumblebee on Spotify

sub_bees <- read_csv(here::here("inputs/data/subset_bumblebee.csv"))

sub_bees |>
  select(artist_name,
         valence,
         danceability,
         album_release_year,
         track_name) |>
  head(17) |>
  knitr::kable(
    col.names = c(
      "Artist Name",
      "Valence",
      "Danceability",
      "Album Release Year",
      "Track Name"
    )
  )
```

@@

```{r}
#| label: fig-quadvalence
#| fig-cap: Valence and Energy Quadrant
#| echo: false
#| warning: false

df <- read_csv(here::here("inputs/data/df.csv"))

ggplot(df, aes(x = energy, y = valence)) +
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$valence), color = "gray") +
  theme_minimal()
```

@@

ggplot(df, aes(x = energy, y = danceability)) +
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$danceability),
             color = "gray") +
  theme_minimal()
```

@@

By comparing cumulative distribution functions of each variable based on whether a song was produced by Antonoff, we can see that some of these variables do capture some sort of underlying Jack-iness (@fig-cdf). For instance, while liveness, instrumentalness, and speechiness are relatively similar, we can see clear differences in danceability, energy, loudness, acousticness, and valence.

```{r}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable, 
#| echo: false
#| warning: false

data_reduced <- read_csv(here::here("inputs/data/data_reduced.csv"))

data_reduced |>
  pivot_longer(
    cols = c(
      danceability,
      energy,
      loudness,
      mode,
      speechiness,
      acousticness,
      instrumentalness,
      liveness,
      valence,
      tempo
    ),
    names_to = "name",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, color = jack)) +
  stat_ecdf() +
  facet_wrap(vars(name),
             nrow = 3,
             ncol = 4,
             scales = "free") +
  theme_minimal() +
  labs(x = "Value",
       y = "Proportion")

#even tho there's similrities, there's differences
```

Another way to gauge the Antonoff-ification of pop music could be to look at how artists' sounds have changed over time, and whether there is any correlation with Antonoff. Here is where the necessary "correlation is not causation" caveat comes in: suggesting that Antonoff is the *cause* for any change in artists' sounds would be an overstep. These plots are here to get a glimpse at pop's evolution over time, and where Antonoff fits in that picture. Each point represents a song.

@@

df <- read_csv(here::here("inputs/data/df.csv"))

df <- df |> 
  mutate(jack = factor(jack))

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = energy, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Energy") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = danceability, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Danceability") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = valence, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Valence") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = acousticness, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Acousticness") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = instrumentalness, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Instrumentalness") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

df |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = tempo, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Tempo") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

Now that we understand that there do seem to be some subtle differences in how variables that are and aren't produced by Jack behave, as seen in our cumulative distribution functions, our quadrants, and some minor trends in artists' sounds over time (specifically, `energy`), we can use a logistic regression to see if a model can a) predict whether a song was produced by Antonoff based on these variables, and b) tell us which variables were more significant in determining the output.

@@

data_reduced <- read_csv(here::here("inputs/data/data_reduced.csv"))

data_reduced |>
  group_by(jack) |>
  summarise(across(-c(tempo, loudness), mean)) |>
  pivot_longer(cols = -jack) |>
  ggplot(aes(value, name, fill = factor(jack))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")
```

Bracketing the major blindspot that these variables provide a single number to represent the dynamics of an entire song, the model has a few other limitations. First of all, this model assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. (The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.)

@@

data_reduced |>
  count(jack) |>
  knitr::kable(col.names = c("Is it Jack?", "Number of Songs"))
```

@@

After some minor pre-processing using the `tidyverse` R package [@citetidyverse]. I followed @citeSilge to build a model that deals with the imbalanced classes.

@@

jack_vip <- read_csv(here::here("inputs/data/jack_vip.csv"))

jack_vip |>
  knitr::kable(col.names = c("Variable",
                             "Importance",
                             "Sign"))
```

@@

jack_vip |>
  group_by(Sign) |>
  slice_max(Importance, n = 15) |>
  ungroup() |>
  ggplot(aes(Importance, fct_reorder(Variable, Importance), fill = Sign)) +
  geom_col() +
  facet_wrap(vars(Sign), scales = "free_y") +
  labs(y = NULL) +
  theme(legend.position = "none")
```

Danceability, with a positive importance score of 4.32, had the highest level of importance in determining whether a song was produced or co-produced by Jack Antonoff. Instrumentalness comes in second at 2.59, with a negative score (meaning that negative instrumentalness is correlated), and valence comes in third with a negative score at 2.02.

@@

df |>
  filter(jack == 1) |>
  arrange(desc(danceability)) |>
  distinct(track_name, .keep_all = TRUE) |>
  select(artist_name, danceability, track_name) |>
  head(20) |>
  knitr::kable(col.names = c("Artist Name",
                             "Danceability",
                             "Track"))
```

@@

According to this model, though, danceability is the most important feature in predicting whether a song was produced by Jack Antonoff. And the accuracy metrics aren't awful -- the accuracy score is 0.743, and the roc_auc (receiver operating characteristic area under the curve) (more fun to say out loud as an abbreviation. Rock Auck), which is a metric better suited for imbalanced datasets, such as our own, is 0.810. This means that although danceability can't show us the inside of a song, something about this score is telling us something about Jack's music.

@@

df |>
  filter(jack == 1) |>
  ggplot(aes(x = artist_name, y = danceability)) +
  geom_boxplot() +
  ggtitle("Boxplot of Danceability by Artist") +
  coord_flip() +
  theme_minimal()
```

df |>
  filter(jack == 1) |>
  count(artist_name) |>
  arrange(desc(n)) |>
  head(10) |>
  ggplot(aes(x = reorder(artist_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Artists with Most Jack = 1 Values", x = "Artist", y = "Count") +
  theme_minimal()

@@

It only makes sense, of course, to hire a bunch of undergrads and get them to tell us whether a few thousand songs are visceral, and then build our own visceral-ness metric, and see if we get the same results -- that danceability is just visceral-ness in a trench coat, and that visceral-ness is the characteristic that helps us best understand what makes Jack, Jack.
